---
title: "Explorative Data Analysis in R"
author: "Thanmai"
date: "2023-11-18"
output: html_document
---
## R Markdown


The tidyverse is a collection of R packages, including ggplot2, dplyr, and tidyr, designed for coherent and efficient data analysis and visualization through a consistent and tidy data manipulation approach. Install it with install.packages("tidyverse") and load it with library(tidyverse).

```{r}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
library(tidyverse)
```

* **dplyr** : A versatile R package for data manipulation that provides intuitive functions like filter, select, group_by, and summarize, facilitating efficient and readable data wrangling.

* **skimr** : A handy R package offering summary statistics and data summaries with the skim() function, providing a quick overview of numeric and categorical variables in a dataset.

* **janitor** : An R package for data cleaning and tidying, providing functions like clean_names() for standardized column names and remove_empty() for handling missing values and empty rows efficiently.

Install and loading the packages

```{r}

install.packages("dplyr",repos = "http://cran.us.r-project.org")
install.packages("skimr",repos = "http://cran.us.r-project.org")
install.packages("janitor",repos = "http://cran.us.r-project.org")

library(dplyr)
library(skimr)
library(janitor)

```
```{r}
install.packages("readr")
library(readr)

```
```{r}
# To get the current directory of this project

getwd()
```
Load the diamonds dataset using **readr** package **read.csv()**. Just print the top 5 rows of dataset using head()

```{r}
library(datasets)
setwd("C://Users/DEVIKA/Desktop/R_Codes/Project")
df = read.csv("diamonds.csv")
head(df,5)
```
**glimpse()** : A function from the **dplyr** package that provides a concise and informative summary of a data frame, displaying variable types, first few observations, and additional metadata.

```{r}
glimpse(df)
```
**str()** : A base R function that displays the structure of an object, providing information about its type and the first few elements or components.

```{r}
str(df)
```
**colnames()** : A base R function used to retrieve or set the column names of a data frame or matrix, facilitating easy manipulation and exploration of variable names.

```{r}
colnames(df)
```
**tibble:** An R package providing the `tibble` class, an enhanced version of data frames. Tibbles are part of the `tidyverse` ecosystem, offering improved printing, subsetting, and compatibility with modern data analysis workflows.

Install and load tibble package
```{r}
install.packages("tibble")
library(tibble)
as_tibble(df)
```

**summary()** : A base R function used to generate summary statistics of numerical data, providing key measures such as mean, median, quartiles, and extreme values for a quick overview of the distribution.

```{r}
summary(df)
```
Here we are looking to get trends in **carat** variable of dataset

```{r}
summary(df$carat)
```
Here we are looking to get trends in **color** variable of dataset

```{r}
summary(df$color)
```
## Data Transforming Techniques

**skim_without_charts()** : A function from the **skimr** package that generates a concise summary of data statistics without charts, offering key insights into the distribution and characteristics of variables in a dataset.

```{r}
library(skimr)
skim_without_charts(df)
```
**select()**: A function from the **dplyr** package used for column selection in data frames, allowing users to choose and reorder specific columns to streamline data manipulation tasks.
In this we are selecting only **cut** column of dataset.

```{r}
library(dplyr)

new_df <- df %>% 
  select(cut)   ## only species
head(new_df,5)
```
**count()** : A function from the **dplyr** package that is used to count the number of observations or occurrences of unique values within a data frame, facilitating basic frequency analysis.
The code chunk gives no of observations of each diamond cut type.

```{r}
diamonds %>% count(cut)
```
The above code chunk selects multiple columns of the dataset i.e trimming the dataset

```{r}
trimmed_df <- df %>% 
  select(carat, cut, price)
head(trimmed_df,5)
```
The code below provides us with summary of cut column.

```{r}
count_cuts <- df %>%
  group_by(cut) %>%
  summarise(count = n()) 
View(count_cuts)
```

**rename()** : A function from the **dplyr** package that enables the renaming of columns in a data frame, providing a flexible and straightforward way to update variable names.

The code renames the cut column to cut_model.
```{r}
renamed <- df %>% 
  rename(cut_model = cut)
head(renamed,5)
```

```{r}
install.packages("janitor")
library(janitor)

#clean_names(df)
```
**unite()** : A function from the **tidyr** package in the **tidyverse** ecosystem, used for combining or uniting multiple columns into a single column by concatenating their values, helping reshape and restructure data efficiently.

A data cleaning function, which combines x,y,z values under measurements column.

```{r}
united_df <- df %>%
  unite(measurements, x, y, z, sep = ", ")
head(united_df,5)
```
**mutate()** : A function from the **dplyr** package that allows the creation of new variables or the modification of existing ones within a data frame, providing a powerful tool for data manipulation and feature engineering.

The above code modifies price column by decreasing by 50 dollars and updating as diwali_sale_price.

```{r}
sale <- united_df %>%
  mutate(diwali_sale_price = price - 50)
head(sale,10)
```
This code removes the original price column of dataset.

```{r}
diamonds <- df %>%
  mutate(diwali_sale_price = price - 50) %>%
  select(-price) 
head(diamonds,10)
```

Some statistical functions to find stats of various variable like minimum, maximum , mean, sum etc.

```{r}

summary_stats <- sale %>%
  summarize(
    total_carat = sum(carat),
    mean_price = mean(price),
    min_depth = min(depth),
    max_table = max(table),
    median_diwali_sale_price = median(diwali_sale_price)
  )

summary_stats
```

```{r}
result <- df %>%
  filter(cut %in% c("Ideal", "Premium"), color %in% c("D", "E", "F")) %>%
  group_by(cut, color) %>%
  summarise(
    avg_price = mean(price),
    total_carat = sum(carat),
    min_depth = min(depth),
    max_table = max(table),
    count = n()
  ) %>%
  arrange(desc(avg_price))

result
```
**filter()** : A function from the **dplyr** package in the **tidyverse** ecosystem, utilized for subsetting or filtering rows of a data frame based on specified conditions, providing a flexible and intuitive approach to data extraction.

**arrange()** : A function from the **dplyr** package in the **tidyverse** ecosystem, used to reorder rows of a data frame based on one or more columns, facilitating easy exploration and analysis of data in a desired order.


The below code filters only **Good** and **premium** values of cut model of Color **E** with depth greater than 60 and arranging in ascending values and then displaying few columns of dataset.

```{r}
filtered <- df %>%
  filter(cut=="Good" | cut=="premium") %>%
  filter(color=="E" & depth>60.0) %>%
  arrange(depth) %>%
  select(carat,cut,color,depth,price)
head(filtered,10)
```

The below code filters the observations having width of diamond 'y' between 3 and 20 using **between()** from **dplyr** package. 

```{r}
diamonds2 <- diamonds %>% 
  filter(between(y, 3, 20))
head(diamonds2,15)
```
**ggplot2:** An R package within the **tidyverse** ecosystem, providing a powerful and flexible system for creating static and dynamic visualizations. `ggplot2` employs a grammar of graphics approach, allowing users to build complex and customized plots layer by layer.

Installing and uploading ggplot2 package.

```{r}
install.packages("ggplot2")

library(ggplot2)
```
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = clarity))
```

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

```{r}
price_summary <- df %>%
  group_by(clarity) %>%
  summarize(median_price = median(price))
price_summary
```

```{r}
library(dplyr)
price_summary <- df %>%
  group_by(clarity) %>%
  summarize(median_price = median(price))

# Bar chart
ggplot(price_summary, aes(x = clarity, y = median_price)) +
  geom_bar(stat = "identity", fill = "lightgreen")
  
```


```{r}
ggplot(data=df) + 
  geom_point(mapping=aes(x=carat,y=price)) 
```

```{r}
ggplot(data = df) + 
  geom_point(mapping = aes(x = carat, y = price), alpha = 0.1)
```

```{r}
ggplot(data=df) +
  geom_smooth(mapping=aes(x=carat ,y=price, linetype = cut, color=cut))
```

```{r}
ggplot(data=df) +
  geom_smooth(mapping=aes(x=carat ,y=price, color=cut))
```

```{r}
ggplot(data=df) +
  geom_smooth(mapping=aes(x=carat ,y=price, color=cut))
```

```{r}
ggplot(data=df) +
  geom_point(mapping=aes(x=carat ,y=price, color=cut, shape = cut))

```

```{r}
ggplot(data=df) +
  geom_point(mapping=aes(x=carat ,y=price, color=cut, shape = cut, size=color))
```

```{r}
ggplot(data=df) +
  geom_smooth(mapping=aes(x=carat,y=price)) +
  geom_point(mapping=aes(x=carat,y=price))
```

```{r}
library(ggplot2)
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.1)
```

```{r}
library(dplyr)
result <- df %>% filter(x == y)
head(result,5)

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)
```
```{r}
boxplot(x = df$X)
```
```{r}
library(ggplot2)
ggplot(df, aes(x = cut, y = price)) +
  geom_boxplot(fill = "skyblue")
```

```{r}
ggplot(diamonds, aes(x = clarity, y = depth)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Boxplot of Diamond Depths by Clarity",subtitle = "Clarity with 8 values", caption = "By Devika", x = "Clarity", y = "Depth")
```

```{r}
library(dplyr)
df %>% 
  count(color, cut) %>%  
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = n))
```

```{r}
ggplot(data=df) +
  geom_point(mapping=aes(x=carat, price)) +
  facet_wrap(~cut)
```

```{r}
ggplot(data=df) +
  geom_point(mapping=aes(x=carat, price)) +
  facet_wrap(~color)
```

```{r}
ggplot(data=df) +
  geom_smooth(mapping=aes(x=carat, y=price)) +
  facet_wrap(~color)
```

```{r}
ggplot(data=df) +
  geom_point(mapping=aes(x=carat,y=price,color=cut)) +
  facet_grid(~color)
```


